\LARGE{\bf Parallelizing a Network Simulator}

### 2009-03-16 doc/paper.tex
title lol
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1154 ce7aba0d-9f81-40a6-a495-cecae84b0515
Fixed figure dependencies after removing multicols.
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1155 ce7aba0d-9f81-40a6-a495-cecae84b0515

\documentclass{article}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage[all]{xy}
\usepackage{multicol}
\usepackage{indentfirst}
\usepackage{algorithm}
\usepackage{algorithmic}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.5in}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\topskip}{0.5in}
\begin{document}
%% be sure to change the:
% Section number, Day, and time
% Date and year
% Name
% spellcheck using ``aspell -c filename'' in Unix
\begin{center}
\LARGE{\bf Parallelizing a Network Simulator}
%% PUT IN YOUR SECTION, DAY, AND TIME
%%example:
%% Section 7 - Thursday 8:30 AM\\

%% PUT IN THE DATE AND YEAR THE ASSIGNMENT IS DUE
%%example:
%% September 10, 2004
%% PUT IN YOUR NAME
\begin{center}
\Large{
Michael Bauer \hspace{1in} Nan Jiang
}
\end{center}
\Large{Department of Computer Science \\ Stanford University \\ \{mebauer, njiang37 \}@stanford.edu}
\end{center}

\begin{multicols}{2}
\begin{center}
\Large{\bf Abstract}
\end{center}
%\small{
\emph{In the context of the current multicore revolution, the need to
design an effective interconnect network has become increasingly
vital.  Computer architects make use of network simulators in order to
determine the best network designs and configurations.  The process by which this
is done however is often computationally intensive and requires
significant run time.  The fact that most simulators are single threaded
only adds to the computational bottleneck and renders them unable to take
advantage of new multicore processors.  This paper demonstrates how
an interconnect network simulator can be parallelized into a multithreaded
implementation.  The additional
performance gain from this parallel simulator ultimately can lead to decreased
development time and better designs as architects can sweep additional
parameter spaces and consider more configurations.}
%}
\section{Introduction}
After hitting the frequency and power walls, chip manufacturers have begun
to design and produce multicore chips that are capable of executing
multithreaded programs.  While this has greatly increased the computation power of the processors, existing programs that have been
written in a single threaded programming model must be rewritten in a
parallel programming model in order to take advantage of the additional cores.  This process of converting a program from a single threaded
programming model to a parallel programming model is a non-trivial task
that often requires restructuring the algorithm of the program as well as
how the data is organized and partitioned.  In this paper we demonstrate
several parallelization techniques in the process of parallelizing a
network simulator.\\
~\\
Another major ramification of the multicore revolution has been the
increased importance of interconnect network design.  As the number of
cores being placed on a chip increases, the processor interconnect plays a bigger and
bigger role in minimizing the latency of communication between processors
and providing enough bandwidth to allow for the movement of large chunks
of data. Traditional bus architecture is no longer the appropriate medium of communication for high number of core. Even more recent single crossbar communication hubs will also not be sufficient in the future, since the complexity of a single crossbar grows quadratically with the number of cores. Interconnection network can provide the scalable and efficient communication
infrastructures for future multicore processors. \\
~\\
In order to make informed decisions in the process of designing
interconnect networks, computer architects make use of simulators to
observe the effect of different network configurations and parameters.  While
these simulations can provide valuable insight into the best designs, they
often require sweeping a large, multi-dimensional parameter space.  In
addition, each simulation requires that many thousands of cycles
of network operation to be simulated.  This demands a large number of computational
resources in order to fully explore the design space.\\
~\\
The network simulator that we employ in this papaer was originally written
in a single threaded programming model.  This made the simulator extremely
slow and a single simulation of just a few thousand cycles would take many
minutes of compute time.  In order to speed up the process of running
simulations we have parallelized this simulator by making it
multithreaded.  In section \ref{background} we provide some background on
the simulator and its mechanics.  In section \ref{para} we describe the
means by which we parallelized the simulator.  We then describe in section
\ref{results} the results that we obtained from our parallel
implementation.  After demonstrating our performance gain and its limiting
factors we discuss how additional performance could be gained in section
\ref{disc}.  Finally, in section \ref{conc} we conclude.

@@

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechansims on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{center}
\includegraphics[width=3in,height=3in]{synch.eps} \\
\small{Figure 1: This figure shows the performance of the different
synchronization mechanisms on up to eight threads.}
\end{center}
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues , we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

\subsection{Determining Scalability Bottlenecks}
Having observed the somewhat poor performance of our application we then
set about attempting to determine the reason for such poor performance.
Our initial hypothesis was that application was memory bound based on the
fact that there is a significant amount of data being processed in our
simulator.  In addition to this, the fact that Pthreads is a shared memory
programming model implies that if the working set cannot fit inside the L2
cache, then there will be a significant amount of data being moved both to
and from main memory.\\
~\\
To test our hypothesis that the simulator was memory bound, we ran the
simulator again on the Hotbox cluster, but this time ran each thread on a
different node.  This allowed each thread to have access to the entire
memory bandwidth of its node, and in effect, thereby increase the total
amount of memory bandwidth available to the entire application.  The
results of this experiment can be seen in Figure \ref{smp}.\\
\begin{center}
\includegraphics[height=3in,width=3in]{smp.eps} \\
\small{Figure 2: This figure shows the performance of the simulator on
up to 16 threads on a simulated SMP machine.  The speedup is relative to
the sequential simulator.}
\end{center}
From the result of this experiment we were able to see that bandwidth was
not the primary issue.  This is due to the fact that the performance of
the simulator remained poor and was even worse than the CMP version up to
16 threads.  In addition to revealing that the poor performance was not
due to memory bandwidth, it also revealed that communication was playing a
large role in the performance as the increased cost of communicating via
ethernet between nodes caused performance to be even worse than the CMP
configuration.\\
~\\
After realizing that our program was not memory bound, we wanted to see if
the size of the problem had any effect on the speedup.  We tested this by
plotting the speedup of a four threaded run on a given network size over the
sequential time for simulating the same size network on the Hotbox
cluster.  The results can be seen in Figure \ref{probsize}.
\begin{center}
\includegraphics[height=3in,width=3in]{probsize.eps}
\small{Figure 3: This figure shows the speedup versus network size of a
four threaded run of the parallel simulator on the Hotbox cluster.}
\end{center}
From this figure we could see that essentially the speedup was independent
of problem size.  This meant that regardless of the size of the network
simulated, the four-threaded version ultimately was constrained by the
same relative to the sequential version.  The fact that the speedup
attained was significantly less than a linear speedup of 4 indicated that
the threads were not compute bound.  Instead we hypothesized that false
sharing was causing the four threads to thrash each others' caches.\\
~\\
To confirm our diagnosis of false sharing, we decided to run the simulator
on the both the Niagara T2 and T2+ machines.  Since these machines were
better designed to deal with long latency loads and stores, our hope was
that they would be able to mitigate the effects of false sharing to some
extend.  The results from this experiment can be seen in figure
\ref{niagara}.
\begin{center}
\includegraphics[height=3in,width=3in]{niagara.eps}
\small{Figure 4: This figure shows the speedup of the parallel simulator
up to 64 threads on the Niagara T2 and T2+ machines.}
\end{center}
We noticed that in this experiment, both the T2 and T2+ were able to
achieve some speedup to a certain number of threads, but then performance
began to fall off.  We determined that once the threads' working set sizes
began to interfere with each other in the shared L2 cache is when the
performance begins to degrade and couldn't offer us much insight into  any potential false sharing that was occurring when run on other
hardware platforms.

@@

\bibliographystyle{IEEEtran}
\bibliography{paper}
\end{multicols}

>>>>

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage[all]{xy}
%\usepackage{multicol}
\usepackage{indentfirst}
\usepackage{algorithm}
\usepackage{algorithmic}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.5in}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\topskip}{0.5in}
%% be sure to change the:
% Section number, Day, and time
% Date and year
% Name
% spellcheck using ``aspell -c filename'' in Unix
%\begin{center}
%\LARGE{\bf Parallelizing a Network Simulator}
%% PUT IN YOUR SECTION, DAY, AND TIME
%%example:
%% Section 7 - Thursday 8:30 AM\\

%% PUT IN THE DATE AND YEAR THE ASSIGNMENT IS DUE
%%example:
%% September 10, 2004
%% PUT IN YOUR NAME
%\begin{center}
%\Large{
%Michael Bauer \hspace{1in} Nan Jiang
%}
%\end{center}
%\Large{Department of Computer Science \\ Stanford University \\ \{mebauer, njiang37 %\}@stanford.edu}
%\end{center}

\title{\LARGE{\bf Parallelizing a Network Simulator}}
\author{
Michael Bauer \hspace{1in} Nan Jiang
}

\date{\Large{Department of Computer Science \\ Stanford University \\ \{mebauer, njiang37 \}@stanford.edu}}
%\maketitle

%\begin{multicols}{2}
%\begin{center}
%\Large{\bf Abstract}
%\end{center}
%\small{
%\emph{
\begin{document}
\maketitle
\abstract{ 
In the context of the current multicore revolution, the need to
design an effective interconnect network has become increasingly
vital.  Computer architects make use of network simulators in order to
determine the best network designs and configurations.  The process by which this
is done however is often computationally intensive and requires
significant run time.  The fact that most simulators are single threaded
only adds to the computational bottleneck and renders them unable to take
advantage of new multicore processors.  This paper demonstrates how
an interconnect network simulator can be parallelized into a multithreaded
implementation.  The additional
performance gain from this parallel simulator ultimately can lead to decreased
development time and better designs as architects can sweep additional
parameter spaces and consider more configurations.}
%}
\section{Introduction}
After hitting the frequency and power walls, chip manufacturers have begun
to design and produce multicore chips that are capable of executing
multithreaded programs.  While this has greatly increased the computation power of the processors, existing programs that have been
written in a single threaded programming model must be rewritten in a
parallel programming model in order to take advantage of the additional cores.  This process of converting a program from a single threaded
programming model to a parallel programming model is a non-trivial task
that often requires restructuring the algorithm of the program as well as
how the data is organized and partitioned.  In this paper we demonstrate
several parallelization techniques in the process of parallelizing a
network simulator.\\
~\\
Another major ramification of the multicore revolution has been the
increased importance of interconnect network design.  As the number of
cores being placed on a chip increases, the processor interconnect plays a bigger and
bigger role in minimizing the latency of communication between processors
and providing enough bandwidth to allow for the movement of large chunks
of data. Traditional bus architecture is no longer the appropriate medium of communication for high number of core. Even more recent single crossbar communication hubs will also not be sufficient in the future, since the complexity of a single crossbar grows quadratically with the number of cores. Interconnection network can provide the scalable and efficient communication
infrastructures for future multicore processors. \\
~\\
In order to make informed decisions in the process of designing
interconnect networks, computer architects make use of simulators to
observe the effect of different network configurations and parameters.  While
these simulations can provide valuable insight into the best designs, they
often require sweeping a large, multi-dimensional parameter space.  In
addition, each simulation requires that many thousands of cycles
of network operation to be simulated.  This demands a large number of computational
resources in order to fully explore the design space.\\
~\\
The network simulator that we employ in this papaer was originally written
in a single threaded programming model.  This made the simulator extremely
slow and a single simulation of just a few thousand cycles would take many
minutes of compute time.  In order to speed up the process of running
simulations we have parallelized this simulator by making it
multithreaded.  In section \ref{background} we provide some background on
the simulator and its mechanics.  In section \ref{para} we describe the
means by which we parallelized the simulator.  We then describe in section
\ref{results} the results that we obtained from our parallel
implementation.  After demonstrating our performance gain and its limiting
factors we discuss how additional performance could be gained in section
\ref{disc}.  Finally, in section \ref{conc} we conclude.

@@

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechansims on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{figure}[h]
\centering
\includegraphics[width=3in,height=3in]{synch.eps}
\caption{\label{org} This figure shows the performance of the different
synchronization mechanisms on up to eight threads.}
\end{figure}
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues , we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

\subsection{Determining Scalability Bottlenecks}
Having observed the somewhat poor performance of our application we then
set about attempting to determine the reason for such poor performance.
Our initial hypothesis was that application was memory bound based on the
fact that there is a significant amount of data being processed in our
simulator.  In addition to this, the fact that Pthreads is a shared memory
programming model implies that if the working set cannot fit inside the L2
cache, then there will be a significant amount of data being moved both to
and from main memory.\\
~\\
To test our hypothesis that the simulator was memory bound, we ran the
simulator again on the Hotbox cluster, but this time ran each thread on a
different node.  This allowed each thread to have access to the entire
memory bandwidth of its node, and in effect, thereby increase the total
amount of memory bandwidth available to the entire application.  The
results of this experiment can be seen in Figure \ref{smp}.\\
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{smp.eps}
\caption{\label{smp}This figure shows the performance of the simulator on
up to 16 threads on a simulated SMP machine.  The speedup is relative to
the sequential simulator.}
\end{figure}
From the result of this experiment we were able to see that bandwidth was
not the primary issue.  This is due to the fact that the performance of
the simulator remained poor and was even worse than the CMP version up to
16 threads.  In addition to revealing that the poor performance was not
due to memory bandwidth, it also revealed that communication was playing a
large role in the performance as the increased cost of communicating via
ethernet between nodes caused performance to be even worse than the CMP
configuration.\\
~\\
After realizing that our program was not memory bound, we wanted to see if
the size of the problem had any effect on the speedup.  We tested this by
plotting the speedup of a four threaded run on a given network size over the
sequential time for simulating the same size network on the Hotbox
cluster.  The results can be seen in Figure \ref{probsize}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{probsize.eps}
\caption{\label{probsize}This figure shows the speedup versus network size of a
four threaded run of the parallel simulator on the Hotbox cluster.}
\end{figure}
From this figure we could see that essentially the speedup was independent
of problem size.  This meant that regardless of the size of the network
simulated, the four-threaded version ultimately was constrained by the
same relative to the sequential version.  The fact that the speedup
attained was significantly less than a linear speedup of 4 indicated that
the threads were not compute bound.  Instead we hypothesized that false
sharing was causing the four threads to thrash each others' caches.\\
~\\
To confirm our diagnosis of false sharing, we decided to run the simulator
on the both the Niagara T2 and T2+ machines.  Since these machines were
better designed to deal with long latency loads and stores, our hope was
that they would be able to mitigate the effects of false sharing to some
extend.  The results from this experiment can be seen in figure
\ref{niagara}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{niagara.eps}
\caption{\label{niagara}This figure shows the speedup of the parallel simulator
up to 64 threads on the Niagara T2 and T2+ machines.}
\end{figure}
We noticed that in this experiment, both the T2 and T2+ were able to
achieve some speedup to a certain number of threads, but then performance
began to fall off.  We determined that once the threads' working set sizes
began to interfere with each other in the shared L2 cache is when the
performance begins to degrade and couldn't offer us much insight into  any potential false sharing that was occurring when run on other
hardware platforms.

@@

\bibliographystyle{IEEEtran}
\bibliography{paper}
%\end{multicols}
